Smart Representations: {R}ationality and Evolution in a Richer Environment

### 2015-12-18 paper/paper.tex
sec com

If agents deal with a rich and variable environment, they face many different choice
situations. A choice mechanism associates decision situations with action choices. A crucial
part of any choice mechanism is a \emph{subjective representation} of the decision situation,
in particular a manner of forming preferences and beliefs about a possibly uncertain world (see
Figure~\ref{fig:ChoiceMechanism} below). This paper asks: which preference and belief
representations are ecologically valuable (i.e., lead to high fitness)? The evolution of
preferences has been subject of recent interest in theoretical economics
\citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. The particular contribution of this paper is
to provide a case, spelled out formally, that questions of preference evolution should take
variability in belief representation into account as well. As a point in case, we demonstrate
that if agents have imprecise probabilistic beliefs \citep[e.g.][]{gardsah82,levi74,walley96},
preference representations in terms of regret that deviate from the true, objective fitness
that evolutionary selection operates on can be selected for.

@@

The fitness of a subjective representation scheme is measured in terms of the fitness of a
choice mechanism that contains it, all else equal. The fitness of a choice mechanism $c$ is the
expected payoff of $c$ over arbitrary decision situations $d$:
\begin{align}
  \label{eq:FittnessChoiceMechSolitary}
  F(c) = \int P(d) \  \utils_d(c) \text{ d} d \,,
\end{align}
where $P(d)$ is the probability (density) of $d$ occurring and $\Utils_d(c)$ i the payoff of
the act selected by $c$ in $d$. Probability measure $P(d)$ is our model of the statistical
properties of a rich and variable environment.

@@

\begin{definition}[Meta game]
  \label{def:MetaGame}
  Let $C$ be a set of choice mechanisms and let $P(g)$ give the occurrence probability of game
  $g$ in the environment. The meta game that captures selection of choice mechanisms in this
  environment is the one-population evolutionary game with strategies $C$ and payoff function
  $\pi \mycolon C \times C \rightarrow \mathbb{R}$ such that
  $\pi(c,c') = F(c,c')$.
\end{definition}

@@

We submit that meta games are a useful tool for structuring and sharpening conceptual inquiry
in a domain, evolutionary justifications, that is notorious for its fuzziness and
elusiveness. To show how this approach can be applied to benefit conceptual debate, we consider
the evolution of preference. Concretely, we here pit two different ways of representing
preferences against each other: (i) veridical representations that correctly reflect objective
fitness, and (ii) non-veridical subjective preferences in terms of regret
\citep[e.g.,][]{Savage1951:The-theory-of-s,}. We show that subjective, non-veridical preference
representations based on a notion of regret can be adaptive and even outperform objective,
veridical representations of the actual, true fitness, when agents, at least on occasion, hold
\emph{imprecise probabilistic beliefs} \citep[e.g.,][]{gilsch89,levi74,gardsah82} and make
decisions based on a security choice rule (here: the maxmin rule applied to expected
utility).

This case study strips the generality of a meta game approach down to a perspicuous comparison
between only four choice mechanisms: the cross-product of two different preference
representations (veridical vs.~regret) and two different belief representations (precise
vs.~imprecise). The following section is dedicated to introduce and justify this exclusive
selection.

@@

\subsection{Classical rules for rational choice} 
\label{sec:choice-rules}

The classical literature on rational choice contains a number of proposals for which acts a
player \emph{should} choose if she is rational. Maximization of expected utilities, as
given in Definition~\ref{def:rationality}, is only one of them. Here, we consider three others:
(i) maxmin, (ii) Laplace's rule and (iii) regret minimization. All of these apply to games and
solitary decision situations. We focus on games, but definitions carry over straightforwardly.

A strategic game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a finite set of
players, $A_i$ a finite set of acts and $\pi_i$ a payoff function for each player $i \in N$.
For the most part, this paper will focus on symmetric games with two players. A two-player game
is symmetric if players have the same action set, $A_1 = A_2$, and payoffs are symmetric:
$\pi_1(a_k, a_j) = \pi_2(a_j, a_k)$ for all $a_j,a_k \in A_1$. An example is given in
Figure~\ref{coordgame1}. The numbers give the payoff for the player choosing a row act against
an opponent choosing a column act. Since the game is symmetric, it suffices to give just these
numbers.

Rational choice from maximization of expected utilities, as given in
Definition~\ref{def:rationality}, applies to games. To say which acts are rational, it requires
specifying a belief of the DM. Let $A_{-i} = \bigtimes A_j$ for all $j \neq i$ be the set of
all \emph{profiles} of acts for all players except $i$. If player $i$ has a belief
$\mu_i \in \Delta(A_{-i})$ about the behavior of his opponents, he chooses rationally only if
his choice $a^*_i$ maximizes expected utility
$\EU(a_i) := \sum_{\act_{-i} \in \Acts_{-i}} \mu_i(\act_{-i}) \myts \pi_i(a_i, a_{-i})$. So
this requires a probabilistic belief about the opponents' action choices. In the example from
Figure~\ref{coordgame1}, a rational player would choose act $I$ exactly when he believes that
his opponent will choose act $I$ with a probability of at least $\nicefrac{2}{3}$.

Perhaps the most well-known choice rule that does not require explicitly stating the DM's
beliefs is the \emph{maxmin rule}. It tells our agent to choose act $\act^*$ only if it
maximizes payoffs for all worst case scenarios. For a strategic game with 2 players this can be
formalized as
$\act^* \in \argmax_{\myact{i} \in \Acts_i} \min_{\myact{j} \in \Acts_j} \pi_i(\myact{i},
\myact{j})$.
For the example in Figure~\ref{coordgame1}, the maximim rule looks at the minima in each row,
asking: what is the worst possible outcome for each row act? It is $0$ in either
case. Consequently, the maxmin rule would not favor either act: players choosing by the
maxmin rule would be indifferent between acts $I$ and $II$.

Another well-known choice rule is \emph{Laplace's rule} which prescribes to choose $\act^*$ if
$\act^* \in \argmax_{a \in A} \sum_{a'} \pi(a, a')$. Laplace's rule is equivalent to expected
utility maximization under a \emph{flat belief} that assigns equal probability to all of the
opponent's choices. In the coordination game of Figure~\ref{coordgame1}, Laplace's rule selects
act $II$.\todo{definitions for maxmin and Laplace are for 2 player games only; rest for
  $n$-players; fix!}

The final choice rule that is relevant for our purposes here is \emph{regret
  minimization}. The notion of regret in decision theory dates back at least to the work by
\citet{Savage1951:The-theory-of-s}, and has later been developed by \citet{bell82}, \citet{fish82} and \citet{loosug82} independently. Lately,
\citet{HalpernPass2012:Iterated-Regret} showed how the use of regret minimization can give
solutions to game theoretical puzzles (like the Traveller's dilemma) in a way that is closer to
everyday intuition and empirical data. In this paper the notion of regret is defined as in
\citet{HalpernPass2012:Iterated-Regret}:

@@

\noindent Acts selected by the maxmin rule for the negative regret transformation of $G$ are
always exactly the acts selected by regret minimization for the original game $G$.\todo{can we
  cite a paper that proves this? should we give a proof ourselves?}

@@

In sum, classical proposals for rational choice are normative prescription of action choices
for a rational agent. Such prescriptions may rest on implicit assumptions about what the DM
believes (e.g., when interpreting Laplace's rule as rational choice under flat beliefs). They may
also make implicit assumptions about what the DM prefers (e.g., when reconstructing regret
minimization in terms of maxmin-choices on the regret transformation). This is rather messy. To
disentangle and avoid confusion, we turn to choice mechanisms.

@@

A \emph{choice mechanism} is a map from choice situations to action choices that, unlike choice
rules, contains an explicit level of subjective representation of a decision situation. For our
purposes, we should think of a choice mechanism as a pair of functions ($\gamma,\kappa$), where
$\gamma$ is a \emph{subjective representation scheme} and $\kappa$ is an \emph{action selection
  function} (see Figure~\ref{fig:ChoiceMechanism}). The subjective representation scheme
$\gamma$ maps a decision situation onto a subjective representation of that decision
situation. The action selection function $\kappa$ maps such a subjective representation onto a
concrete action choice.

% The crucial bit for our purposes here is the representation function
% $\gamma$, because we are interested in studying which ways of conceptualizing a decision
% situation are evolutionarily successful.

Formally, given a class of strategic games $\mathcal{G}$, the subjective representation scheme
for agent $i$ is a function $\gamma_i: \mathcal{G} \rightarrow \Theta \times E$ that gives a
subjective preference $\theta \in \Theta$ and an epistemic state $e \in E$ for any game
$G \in \mathcal{G}$. A subjective preference $\theta$ for $G$ and $i$ is just a utility
function for $G$. This can but need not be different from the actual payoff function of $G$. An
epistemic state $e$ for $G$ and $i$ is some specification of $i$'s uncertainty about the
co-players' behavior. Finally, an action selection function for agent $i$ is a function
$\kappa_i: \Theta \times E \rightarrow \pow{A_i} $, that associates any subjective
representation $\gamma_i(G) = (\theta,e)$ of $G$ with a set of prescribed actions
for $i$ in $G$ under $i$'s subjective representation of $G$. We drop the index $i$ for
readability in the following.

@@

% The following subsection describes what these belief types are and how a single action
% selection function (maxmopin expected utility) gives rise to an intersting parallel with
% classical choice rules.

@@

\subsection{Choice mechanisms \& classical choice rules}

@@

\begin{table}[t]
  \centering
  \begin{tabular}{ccc}
    preference type & belief type & corresponding choice rule \\ \midrule
    $\tau^\pi$ & $\overline{\mu}$ & Laplace \\
    $\tau^\pi$ & $\Delta$ & maxmin \\ 
    $\tau^{reg}$ & $\overline{\mu}$ & Laplace \\
    $\tau^{reg}$ & $\Delta$ & regret minimization \\ 
  \end{tabular}
  \caption{Overview relevant choice mechanisms and the corresponding (behaviorally equivalent) classical choice rule}
  \label{tab:CMs}
\end{table}

@@

\item We argue that this ``meta-evolutionary'' approach could eventually help explain attested
  deviations of human decision making from the classical ideal: choice behavior evolved to be
  an ``ecologically rational'', near-optimal adaptation to the environment
  \citep{Anderson1990:The-Adaptive-Ch,Anderson1991:Is-human-cognit,GigerenzerGoldstein1996:Reasoning-the-F,ChaterOaksford2000:The-Rational-An}. This
  is inline with recent aspirations to provide an evolutionary rationale for decision making
  processes
  \citep[e.g.,][]{HammersteinStevens2012:Six-Reasons-for,FawcettHamblin2013:Exposing-the-be}. Here,
  we seek to contribute to the formal machinery, an extension of evolutionary game theory, that
  allows to capture statistical properties of the environment
  \citep[cf.][]{McNamara2013:Towards-a-Riche}.
\end{itemize}

@@

    attested deviations of human decision making from the classical ideal: choice behavior

>>>>

If agents deal with a rich and variable environment, they face many different choice
situations. A choice mechanism associates decision situations with action choices. A crucial
part of any choice mechanism is a \emph{subjective representation} of the decision situation,
in particular a manner of forming preferences and beliefs about a possibly uncertain world (see
Figure~\ref{fig:ChoiceMechanism} below). This paper asks: which preference and belief
representations are ecologically valuable (i.e., lead to high fitness)? The evolution of
preferences has been subject of recent interest in theoretical economics
\citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. The particular contribution of this paper is
to provide a case, spelled out formally, that questions of preference evolution should take
variability in belief representation into account as well. As a point in case, we demonstrate
that if agents have imprecise probabilistic beliefs \citep[e.g.][]{gardsah82,levi74,walley96},
preference representations in terms of regret can be selected for, even though they may deviate
from the true, objective fitness that evolutionary selection operates on.

@@

The fitness of a subjective representation scheme is measured in terms of the fitness of a
choice mechanism that contains it, all else equal. The fitness of a choice mechanism $c$ is the
expected payoff of $c$ over arbitrary decision situations $d$:
\begin{align}
  \label{eq:FittnessChoiceMechSolitary}
  F(c) = \int P(d) \  \utils_d(c) \text{ d} d \,,
\end{align}
where $P(d)$ is the probability (density) of $d$ occurring and $\Utils_d(c)$ is the payoff of
the act selected by $c$ in $d$. Probability measure $P(d)$ is our model of the statistical
properties of a rich and variable environment.

@@

\begin{definition}[Meta game]
  \label{def:MetaGame}
  Let $C$ be a set of choice mechanisms and let $P(g)$ give the occurrence probability of game
  $g$ in the environment. The meta game that captures selection of choice mechanisms in this
  environment is the one-population evolutionary game with strategies $C$ and payoff function
  $\pi \mycolon C \times C \rightarrow \mathbb{R}$ such that
  $\pi(c,c') = F(c,c') = \int P(g) \  \utils_g(c,c') \text{ d} g$.
\end{definition}

@@

@@

We submit that meta games are a useful tool for structuring and sharpening conceptual inquiry
in a domain that is notorious for its fuzziness and elusiveness. To show how this approach can
be applied to the benefit of conceptual debate, we consider the evolution of preference
\citep[e.g.,][]{algweib13,DekElyYlan07,RobSam11}. Concretely, we here pit two different ways of
representing preferences against each other: (i) veridical representations that correctly
reflect objective fitness, and (ii) non-veridical subjective preferences in terms of regret
\citep[e.g.,][]{Savage1951:The-theory-of-s,}. We show that subjective, non-veridical preference
representations based on a notion of regret can be adaptive and even outperform objective,
veridical representations of the actual, true fitness, when agents, at least on occasion, hold
\emph{imprecise probabilistic beliefs} \citep[e.g.,][]{gilsch89,levi74,gardsah82} and make
decisions based on a security choice rule (here: the maxmin rule applied to expected
utility). This case study strips the generality of a meta game approach down to a perspicuous
comparison between only four choice mechanisms, summarized in Table~\ref{tab:CMs}: the
cross-product of two different preference representations (veridical vs.~regret) and two
different belief representations (precise vs.~imprecise).

\begin{table}[t]
  \centering
  \begin{tabular}{cccc}
    \multicolumn{2}{c}{subjective representation} & action selection & corresponding choice
    rule \\  \cmidrule(r){1-2}
    preference type & belief type  \\ \midrule
    objective fitness & precise & maxmin EU &  Laplace \\
    objective fitness & precise & maxmin EU & maxmin \\ 
    regret & imprecise & maxmin EU & Laplace \\
    regret & imprecise & maxmin EU & regret minimization \\ 
  \end{tabular}
  \caption{Overview relevant choice mechanisms and the corresponding (behaviorally equivalent)
    classic choice rules. All notions are explained in the main text.}
  \label{tab:CMs}
\end{table}

This exclusive selection of choice mechanisms is motivated by a number of
considerations. Firstly, the set of conceivable choice mechanisms is vast. Most of that
conceivable vastness is utter nonsense. Some selection of plausible candidates is required in
any case. Secondly, a severe restriction to only a small number of suitable candidates will
help demonstrate more distinctly how a meta game approach may help structure philosophical
inquiry. This is our main goal. Thirdly, our case study adds to the recent literature on the
evolution of preferences by clearly demonstrating the possibly non-trivial interaction between
preference and belief representations, all else equal. This is our secondary goal. Finally, the
four choice mechanism that we look at here, although phrased in terms of preference and belief
representations, map onto well-known classic proposals for rational choice (see
Table~\ref{tab:CMs}). In other words, there is also an argument by historical relevance for our
selection. To see this, we first recapitulate the relevant classics ---to have a handy label,
we call them \emph{classic rules}--- and then show how they relate to our notion of choice
mechanism.

\subsection{Classic rules for rational choice} 
\label{sec:choice-rules}

The classic literature on rational choice contains a number of proposals for which acts a
player should choose if she is rational. Maximization of expected utilities, as
given in Definition~\ref{def:rationality}, is only one of them. Here, we consider three others:
(i) maxmin, (ii) Laplace's rule and (iii) regret minimization. All of these apply to games and
solitary decision situations. We focus on games, but definitions carry over straightforwardly.

A strategic game is a tuple $ G=\tuple{ N, (A_i , \pi_i)_{i \in N} }$ with $N$ a finite set of
players, $A_i$ a finite set of acts and $\pi_i$ a payoff function for each player $i \in N$.
Part of this paper will focus on symmetric games with two players. A two-player game is
symmetric if players have the same action set, $A_1 = A_2$, and payoffs are symmetric:
$\pi_1(a_k, a_j) = \pi_2(a_j, a_k)$ for all $a_j,a_k \in A_1$. An example is given in
Figure~\ref{coordgame1}. The numbers give the payoff for the player choosing a row act against
an opponent choosing a column act. Since the game is symmetric, it suffices to give just these
numbers.

Rational choice from maximization of expected utilities, as given in
Definition~\ref{def:rationality}, applies to games as follows. To say which acts are rational,
it requires specifying a belief of the DM. Let $A_{-i} = \bigtimes A_j$ for all $j \neq i$ be
the set of all \emph{profiles} of acts for all players except $i$. If player $i$ has a belief
$\mu_i \in \Delta(A_{-i})$ about the behavior of his opponents, he chooses rationally only if
his choice $a^*_i$ maximizes expected utility
$\EU(a_i) := \sum_{\act_{-i} \in \Acts_{-i}} \mu_i(\act_{-i}) \myts \pi_i(a_i, a_{-i})$. For
the example in Figure~\ref{coordgame1}, a rational player would choose act $I$ exactly when he
believes that his opponent will choose act $I$ with a probability of at least
$\nicefrac{2}{3}$.

Perhaps the most well-known classic rule that does not require explicitly stating the DM's
beliefs is the \emph{maxmin rule}. It tells our agent to choose act $\act^*$ only if it
maximizes payoffs for all worst case scenarios:
$\act^* \in \argmax_{\myact{i} \in \Acts_i} \min_{\myact{-i} \in \Acts_{-1}} \pi_i(\myact{i},
\myact{-1})$.
For the example in Figure~\ref{coordgame1}, the maximim rule looks at the minima in each row,
asking: what is the worst possible outcome for each row act? It is $0$ in either
case. Consequently, the maxmin rule would not favor either act: players choosing by the maxmin
rule would be indifferent between acts $I$ and $II$.

\emph{Laplace's rule} selects $\act^*$ if
$\act^* \in \argmax_{\myact{-i} \in \Acts_{-i}} \sum_{\myact{-i}} \pi(\myact{i}, \myact{-i})$.
Laplace's rule is equivalent to expected utility maximization under a \emph{flat belief} that
assigns equal probability to all of the opponents' choices. In the coordination game of
Figure~\ref{coordgame1}, Laplace's rule selects act $II$.

The final classic rule that is relevant for our purposes here is \emph{regret
  minimization}. The notion of regret in decision theory dates back at least to the work by
\citet{Savage1951:The-theory-of-s}, and has later been developed by \citet{bell82}, \citet{fish82} and \citet{loosug82} independently. Lately,
\citet{HalpernPass2012:Iterated-Regret} showed how the use of regret minimization can give
solutions to game theoretical puzzles (like the Traveller's dilemma) in a way that is closer to
everyday intuition and empirical data. In this paper the notion of regret is defined as in
\citet{HalpernPass2012:Iterated-Regret}:

@@

\noindent Acts selected by the maxmin rule for the negative regret transformation of $G$ are
always exactly the acts selected by regret minimization for the original game $G$.

@@

In sum, classic proposals for rational choice are normative prescription of action choices
for a rational agent. Such prescriptions may rest on implicit assumptions about what the DM
believes (e.g., when interpreting Laplace's rule as rational choice under flat beliefs). They may
also make implicit assumptions about what the DM prefers (e.g., when reconstructing regret
minimization in terms of maxmin-choices on the regret transformation). This is rather messy. To
disentangle and avoid confusion, we turn to choice mechanisms.

@@

A \emph{choice mechanism} is a map from choice situations to action choices that, unlike the
classic rules visited above, contains an explicit level of subjective representation of a
decision situation. Let us think of a choice mechanism as a pair of functions
($\gamma,\kappa$), where $\gamma$ is a \emph{subjective representation scheme} and $\kappa$ is
an \emph{action selection function} (see Figure~\ref{fig:ChoiceMechanism}).

The subjective representation scheme $\gamma_i$ of agent $i$ is a function $\gamma_i$ that
takes a strategic game $G = \tuple{ N, (A_i , \pi_i)_{i \in N} }$ and maps it onto
$\gamma_i(G) = (\theta, e_i)$ where
$\theta \mycolon \bigtimes_{i \in N} A_i \rightarrow \mathbb{R}$ is a (subjective) payoff
function for agent $i$ suitable for $G$.

$\mathcal{G} \rightarrow \Theta \times E$, where $\mathcal{G}$ is a class of

Formally, given a class of strategic games $\mathcal{G}$, the subjective representation scheme
for agent $i$ is a function $\gamma_i: \mathcal{G} \rightarrow \Theta \times E$ that gives a
subjective preference $\theta \in \Theta$ and an epistemic state $e \in E$ for any game
$G \in \mathcal{G}$. A subjective preference $\theta$ for $G$ and $i$ is just a utility
function for $G$. This can but need not be different from the actual payoff function of $G$. An
epistemic state $e$ for $G$ and $i$ is some specification of $i$'s uncertainty about the
co-players' behavior.

The action selection function $\kappa$ maps such a subjective representation onto a concrete
action choice. An action selection function for agent $i$ is a function
$\kappa_i: \Theta \times E \rightarrow \pow{A_i} $, that associates any subjective
representation $\gamma_i(G) = (\theta,e)$ of $G$ with a set of prescribed actions for $i$ in
$G$ under $i$'s subjective representation of $G$. We drop the index $i$ for readability in the
following.

@@

% The following subsection describes what these belief types are and how a single action
% selection function (maxmopin expected utility) gives rise to an intersting parallel with
% classic choice rules.

@@

\subsection{Choice mechanisms \& classic choice rules}

@@

@@

\item We argue that this ``meta-evolutionary'' approach could eventually help explain attested
  deviations of human decision making from the classic ideal: choice behavior evolved to be
  an ``ecologically rational'', near-optimal adaptation to the environment
  \citep{Anderson1990:The-Adaptive-Ch,Anderson1991:Is-human-cognit,GigerenzerGoldstein1996:Reasoning-the-F,ChaterOaksford2000:The-Rational-An}. This
  is inline with recent aspirations to provide an evolutionary rationale for decision making
  processes
  \citep[e.g.,][]{HammersteinStevens2012:Six-Reasons-for,FawcettHamblin2013:Exposing-the-be}. Here,
  we seek to contribute to the formal machinery, an extension of evolutionary game theory, that
  allows to capture statistical properties of the environment
  \citep[cf.][]{McNamara2013:Towards-a-Riche}.
\end{itemize}

@@

    attested deviations of human decision making from the classic ideal: choice behavior
